Relation of severe COVID-19 in Scotland to transmission-related factors and risk conditions eligible for shielding support: REACT-SCOT case-control study

### 2021-02-18 shielding.Rmd
updated for shielding paper

---
title: "Relation of severe COVID-19 in Scotland to transmission-related factors in individuals with and without risk conditions eligible for shielding support: REACT-SCOT case-control study"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sam Colville
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    email: helen.colhoun@igmm.ed.ac.uk (+44 7791 652665)
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
#    number_sections: false
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan

@@

* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent hospital exposure -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.

@@

* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .   In all risk groups severe COVID-19 was strongly associated with recent hospital exposure: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall. The fraction of severe cases attributable to recent hospital exposure peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  at the beginning of December 2020. Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.  The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.

* _Conclusions_ -  Mitigating the impact of the epidemic requires control of transmission in hospital settings.  Shielding of vulnerable individuals until they have been vaccinated requires protection from transmission from other adults in the same household.  For solid organ transplant recipients, who are at very high risk and in whom the efficacy of vaccines is uncertain, vaccination should be offered to household contacts.

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].  Those with "very high risk" conditions were designated as clinically extremely vulnerable and eligible for shielding.

The shielding programme issued individual letters to eligible persons advising on the need for social distancing, recommending strict isolation and giving information on the provided support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots. Whether the programme had an impact on the infection rates in those shielded is unclear.

The initial objective of this study was to investigate  whether the shielding letter and programme had any demonstrable impact on the risk of COVID-19 in those eligible for shielding  versus those with moderate risk or the rest of the population.  For this we hypothesized that  a fall in  the risk ratio  for those eligible for  shielding  after the receipt of shielding letters  would be evidence of impact. Our second objective was to understand the role of  transmission-related factors: household composition, recent hospital exposure including out-patient consultations, and exposure related to occupations including health care [@shah_risk_2020] and teaching [@mcallister_report_2020]  on COVID-19 among those eligible for shielding.    As the relevance of these transmission-related factors among the shielded  became clear, the objectives were broadened to investigation of the relation of severe COVID-19 to these transmission-related factors in  the remainder of the population also.

# Methods
In brief; we linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) conducted by PHS that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020-1]   This case control dataset is refreshed every few weeks and  through  linkage to health and other records already contained extensive covariate data on a wide range of risk factors and settings for COVID-19 and derived variables for   "moderate risk conditions". Thus all cases and controls were  classifiable into three categories: no risk condition, moderate risk condition only, eligible for shielding.  The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.

@@

1. Solid organ transplant recipients
2. People with cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
3. People with severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
4. People with rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
5. People on immunosuppression therapies sufficient to increase risk of infection
6. Women who were pregnant with heart disease
7. Other patients identified by clinicians as requiring shielding advice. 
8. Those on renal dialysis and those who had had a splenectomy (included from 11 May 2020)

The dataset of  those eligible for shielding process  has been regularly refreshed.  Here we used the dataset comprising  `r nrow(shielded.full)` individuals identified up to `r dateformat(lastdate.specimen)`. The first batch of  shielding letters was sent on `r dateformat(batches$Date.Sent)[1]`.   Further batches were issued on a weekly basis and the programme was paused on  1 August 2020. On 25 November 2020 a further letter was issued with "extra protection level advice for people at highest risk" based on the current protection level for the population level in that area.

## Ascertainment of cases and sampling of controls
Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020-1].  Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS) that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general hospital discharges including day cases and is ICD-10 coded. All these databases use the Community Health Index (CHI) number as identifier.  The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.

@@

Severe COVID-19 was defined by entry to critical care within 21 days of presentation date, or fatal outcome. Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.  This definition of severe cases was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.

@@

## Linkage of cases and controls to demographic and morbidity data
Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020-1].  Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.  We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.  The codes used are as described previously  [@mckeigue_rapid_2020-1].  Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.  Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].  Quintile 5 is the least deprived.

# Transmission-related risk factors 
Through linkage of the CHI database to the Unique Property Reference Number it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.  Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.  Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].

We used the Scottish Morbidity Records SMR01 and SMR00 (outpatient attendance) to examine possible associations with exposure to health-care facilities. We defined the variable "recent hospital exposure" as any hospital in-patient stay or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all cases that *could have been* nosocomial rather than the more restrictive definition  used in Scotland and England for "probable healthcare -associated infection" as illness onset (or first positive specimen date at least 8 days after admission  which aims to  identify cases that are unlikely to have been infected outside hospital [@public_health_england_covid-19_2020-1].

## Statistical methods
The analyses presented here focus  mainly on severe COVID-19 as this case definition does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is also likely to differ between those eligible for shielding and those ineligible.  Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.

@@

# Results
A methodological concern was that unrecorded filtering of fatalities while the list of those eligible for shielding was being prepared could have led us to misclassify early fatal cases among those eligible for shielding as ineligible, leading to underestimation of the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding. As described in the Supplementary Material, by examining the time course of the fatal case proportion we were able to establish that such misclassification was unlikely in cases presenting after the second week of March 2020.

## Shielding eligibility and risk of COVID-19
Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.  To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women who were identified as in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group.

The largest risk category was those with severe respiratory disease.   Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories. Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters on April 3rd 2020, but most of those in other risk categories were included in later batches.

Table \ref{tab:shieldcasegroup} shows  among the shielding group the number in each eligibility category by case status classified as not diagnosed, not severe or severe COVID-19.

Overall among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome. Those on immunosuppressants had the lowest risks.

Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as a ribbon plot.  The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.   After each peak, the number of daily cases fell more slowly in those eligible for shielding than in those without risk conditions.

## Rate ratio for severe COVID-19  by risk condition group   
Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.  In all subsequent case-control analyses care home residents and their matched controls have been excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.

@@

### Time course of the rate ratio associated with eligibility for shielding 
Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 n the  shielding and moderate risk categories, with those without risk conditions as reference category.   These rate ratios are computed over `r winsize`-day windows of presentation dates.  The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.  Therefore there was no evidence that  the  shielding letters reduced the rate ratio associated with shielding eligibility.  A similar trend of rate ratio over time was found for those with moderate risk conditions.

>>>>

---
title: "Relation of severe COVID-19 in Scotland to transmission-related factors and risk conditions eligible for shielding support: REACT-SCOT case-control study"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: David Caldwell 
    affiliation: HPS
  - name: Ciara Gribben
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Matthew Armstrong
    affiliation: HPS
  - name: Joke Delvaux
    affiliation: HPS
  - name: Sam Colville
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    email: helen.colhoun@igmm.ed.ac.uk (+44 7791 652665)
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
  \linenumbers 
output:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
#    number_sections: false
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./biomed-central.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan

@@

* _Objectives_ - To investigate the relation of transmission-related factors-- occupation, household composition and recent exposure to hospital -- to severe COVID-19 in  clinically vulnerable individuals eligible for shielding and in the general population.

@@

* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was
`r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  The highest risk for severe disease was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])` .   Risk of severe COVID-19 increased with the number of adults in the household but decreased with the number of school-age children in the household.  The rate ratio associated with two or more adults (with single-adult households as reference category), was   `r format.estcipv(table.listedgr3.severecases.nocare["adultsgt1", 5], table.listedgr3.severecases.nocare["adultsgt1", 6])` and the rate ratio associated with one or more school-age children was `r format.estcipv(table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 5], table.listedgr3.severecases.nocare.anyschoolage["hh.schoolage.any", 6])`.  In all risk groups severe COVID-19 was strongly associated with recent exposure to hospital (5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall. In a case-crossover analysis with less recent exposure (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`. The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% of severe cases at the beginning of May 2020, and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  of severe cases at the beginning of December 2020.

* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.  For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts.  <!---These results also provide compelling evidence for a substantial contribution of nosocomial infection to the burden of severe COVID-19---> Mitigating the impact of the epidemic requires control of transmission in hospital settings.

# Background  
The SARS-CoV-2 pandemic reached Scotland in early 2020 with the first positive laboratory test recorded on 1 March 2020. Population-wide interventions included advice that symptomatic individuals should self-isolate, banning of mass gatherings, closure of schools and finally a population-wide lockdown on 23 March 2020.  Although no systematic studies of risk factors were available at that time, public health agencies issued lists of "moderate risk" conditions [@nhs_whos_2020] and "diseases and conditions considered to be very high risk" [@kanani_caring_2020].  Those with "very high risk" conditions were designated as eligible for shielding, and were sent letters recommending strict isolation and offering support which included a national opt-in scheme of free food delivery, home delivery of medication and priority access to supermarket delivery slots.

The initial objective of this study was to investigate  whether shielding advice and support had reduced the risk of COVID-19.  On the hypothesis that shielding was effective, we predicted that the rate ratio associated with eligibility for  shielding  compared to those without risk conditions would fall after the receipt of shielding letters. Our second objective was to understand the role of  transmission-related factors: household composition, occupations including health care [@shah_risk_2020], and recent exposure to hospital on COVID-19 among those eligible for shielding.  As the relevance of these transmission-related factors among those eligible for shielding  became clear, the objectives were broadened to investigate the relation of severe COVID-19 to these transmission-related factors in  the general population.

# Methods
We linked a dataset of all those eligible for shielding  nationally to an existing case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records.  These linked records are used to derive a variables for a list of designated "moderate risk conditions". Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.  The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio  from the case-control study as described below.

@@

1. Solid organ transplant recipient
2. Cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
3. Severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
5. People on immunosuppression therapies sufficient to increase risk of infection
6. Pregnant with heart disease
7. Additional conditions, including people on renal dialysis, those who had had a splenectomy  and others identified by clinicians as requiring shielding advice.

The first batch of  shielding letters was sent on `r dateformat(batches$Date.Sent)[1]`.   Further batches were issued on a weekly basis and the programme was paused on  1 August 2020. On 25 November 2020 a further letter was issued with "extra protection level advice for people at highest risk" based on the current protection level for the population level in that area.   The list of those eligible for shielding has been regularly updated: this study is based on the list of `r nrow(shielded.full)` individuals identified up to 28 January 2021.

## Ascertainment of cases and sampling of controls
Case ascertainment for the REACT-SCOT study has been described in detail elsewhere [@mckeigue_rapid_2020].  Case ascertainment was based on querying the following national-level databases: Electronic Communication of Surveillance in Scotland (ECOSS) that captures virology testing in all NHS laboratories, National Records of Scotland (NRS) death registrations,  RAPID which is a daily update of hospitalisations, and Scottish Morbidity Record 01 (SMR01) which records general hospital discharges including day cases and is ICD-10 coded. All these databases use the Community Health Index (CHI) number as identifier.  The CHI database includes age, sex, postcode and care home status, and can be queried to extract numbers of adults and children in the household.

@@

Entry to critical care units -- intensive care, high dependency or combined units -- was obtained by linkage to the Scottish Intensive Care Society and Audit Group (SICSAG) database. Fatal outcome was defined as death at any time with COVID-19 coded as underlying cause, death from any cause within 28 days of testing positive or death within 28 days of presentation date for cases ascertained through discharge records.  Severe COVID-19 was defined as entry to critical care within 21 days of presentation date, or fatal outcome. This definition was chosen to ensure that case ascertainment would not be affected by triage of those assessed as unlikely to benefit from critical care, or by the scale-up of SARS-CoV-2 testing.

@@

## Linkage of cases and controls to demographic and morbidity data
Linkage of cases and controls to demographic and morbidity data and the associations of these factors with severe COVID-19 have been described in detail elsewhere [@mckeigue_rapid_2020].  Cases and controls were linked to hospital discharge ICD-10 codes over the last five years in SMR01, to British National Formulary codes of dispensed prescriptions in the 240 days before presentation date in the Prescribing Information System (PIS) and to the national register of diabetes.  We used these linked datasets to assign a list of "moderate risk conditions" for COVID-19 designated by public health agencies [@nhs_whos_2020]: diabetes, heart disease, asthma or chronic airway disease, chronic kidney disease, disabling neurological conditions, and immune deficiency or suppression.  The codes used are as described previously  [@mckeigue_rapid_2020].  Three risk groups were defined: no risk condition, moderate risk condition but ineligible for shielding, and eligible for shielding.  Socioeconomic status was assigned as the Scottish Index of Multiple Deprivation (SIMD) score which is based on linkage of postcodes to Census data [@fraser_scottish_2020].  Quintile 5 is the least deprived.

# Transmission-related risk factors 
Since addresses in the CHI database have been mapped to  Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.  Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.  Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].

We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals. We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all cases that *could have been* nosocomial rather than the more stringent definitions for "health-care associated COVID-19" specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].  The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.  Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated. The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g. re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.

## Statistical methods
The analyses presented here focus on severe COVID-19, defined as the main outcome measure of  of the REACT-SCOT study at the design stage.  Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.  Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.

@@

To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated rate ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category. This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases. It is a variant of the classic case-crossover design, which usually  compares in cases only the frequencies of exposure in recent and less recent time windows. Such an analysis can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000]. The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only. The analysis that we report is a refinement of this that takes advantage of the availability of a matched control group.  We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with less recent exposure only. This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.

# Results
## Shielding eligibility and risk of COVID-19
Table \ref{tab:shieldgroupsbyage} shows the frequency of risk categories in those eligible for shielding.  To comply with statistical disclosure control rules, the `r shielded.full[shield.group=="Pregnant with heart disease", .N]` women in the category "pregnant with heart disease" have been omitted from this table and all subsequent analyses; there were no severe cases in this group. The largest risk category was those with severe respiratory disease.   Solid organ transplant recipients and those on immunosuppressants were younger than those in the other categories. Supplementary Table \ref{tab:letterdates} shows the proportions in each of these risk categories by date of advice letter: most of the solid organ transplant recipients, those with severe respiratory disease, and cancer patients were included in the first batch of letters sent on 3 April 2020. 
Table \ref{tab:shieldcasegroup} shows the number in each category of eligibility for shielding by case status classified as not diagnosed, not severe or severe COVID-19; among `r nrow(shielded.full)` persons eligible  for shielding `r table.shield.casegr["All shielding categories", 2]` were diagnosed with COVID-19 but did not enter critical care or die within 28 days, `r table.shield.casegr["All shielding categories", 3]` entered critical care for COVID-19 but survived, and `r table.shield.casegr["All shielding categories", 4]` had a fatal outcome.

Figure \ref{fig:plotcasesbydate} compares the time series of severe cases in the three risk groups as an area plot.  The daily number of presentations of severe cases in the total population reached a peak on `r dateformat(peakdate)` and fell to one quarter of the peak by `r dateformat(deaths.quartermax.date)`.   After the peak, the daily number of severe cases fell more slowly in those eligible for shielding than in those without risk conditions.

## Rate ratio for severe COVID-19  by risk condition group   
Of those eligible for shielding who were included in the case-control dataset, `r cc.severe[CASE==1 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==1 & shield.any==TRUE, .N]` severe cases and `r cc.severe[CASE==0 & shield.any==TRUE, .N, by=care.home][care.home=="Care/nursing home", N]` of the `r cc.severe[CASE==0 & shield.any==TRUE, .N]` matched controls were resident in care homes.  In  subsequent case-control analyses care home residents and their matched controls were excluded, as policies for shielding residents of care homes are different from those relevant to shielding individuals living independently.

@@

### Time course of the rate ratio associated with eligibility for shielding 
Figure \ref{fig:plotrateratio} (upper panel) shows the time course of the univariate rate ratios for severe COVID-19 in the  shielding and moderate risk categories, with those without risk conditions as reference category.   These rate ratios are computed over `r winsize`-day windows of presentation dates.  The rate ratio associated with eligibility for shielding increased from `r round(coeffs.long[date.midpoint==as.Date("2020-04-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 April to `r round(coeffs.long[date.midpoint==as.Date("2020-05-01") & riskgroup=="Eligible for shielding", rateratio], 2)` in the time window with mid-point 1 May.  The rate ratio associated with moderate risk conditions also rose in the first half of May, but fell more rapidly than the rate ratio associated with eligibility for shielding.

A methodological concern was that unrecorded removal of fatalities from the original list of those identified as eligible for shielding could have led us to misclassify early fatal cases among those eligible for shielding as ineligible and thus to underestimate the rate of fatal disease in the first few weeks of the epidemic in those eligible for shielding. As described in Supplementary Methods and Figure \ref{fig:plotcasefatality}, by examining the time course of the fatal case proportion we were able to establish that such misclassification of eligibility was unlikely in cases presenting after the second week of March 2020.

## Associations of severe COVID-19 with transmission-related factors
### Associations in those eligible for shielding 
Supplementary Table \ref{tab:shieldedonlyregressionseverecasesnocare} shows associations with risk factors among those eligible for shielding only, with severe respiratory disease (the largest category) as reference category.  The rate ratio increased with the number of adults in the household but not with the number of children.  The strongest risk factor for severe COVID-19 among those eligible for shielding was recent exposure to hospital, with a rate ratio of `r format.estcipv(table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 5], table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 6])` in the multivariable model.  Of severe cases among those eligible for shielding, `r table.shieldedonly.severecases.nocare["Recent hospital visit/stay", 2]` had recent  exposure to hospital. Using Miettinen's formula as given above, the Population Attributable Risk Fraction  for recent  exposure to hospital can thus be calculated as `r round(100 * pc * (r -1) / r)`% of severe cases in those eligible for shielding.  
<!--Table \ref{tab:casegr} shows the relationship of hospitalisation, entry to critical care and fatality to recent hospital exposure among all diagnosed cases.-->
